---
title: "Home Credit EDA"
author: "Bryson Burr"
date: "February 11, 2025"
output:
  html_document:
    number_sections: true
    toc: true
  pdf_document:
    toc: true
editor_options:
  chunk_output_type: console
---

# Introduction

## Project Goal
The goal of this project is to predict how capable clients are of repaying a loan 
based on application, demographic, and historical credit behavior. This notebook
is to initially explore the given data, trying to find interesting relationships, data problems
or anything that may help in my analytical approach.

## Business Problem Statement
Many people who have insufficient or non-existent credit histories struggle to get loans. Many clients have the ability to repay their loans but get rejected due to their credit histories. This is a problem as failing to accurately predict repayment ability means rejecting customers who could reliably repay their loans. 

## Analytic Problem
Customer repayment probability will be generated using a supervised machine-learning model. The model will use as inputs data collected on past customers such as application, demographic, and historical credit behavior. The model will use this information to predict the probability that the customers will either repay or not repay their loans. 

## Questions to guide exploration
Is the target variable balanced?

How many missing data is there in the dataset? In each column? What do I do with them?

What do I think would be some of the most important factors to determine if someone will pay off their loan?

What are the distributions/relationship with loan repayment for key variables such as age,
income, and employment history?

Are there any highly correlated variables?

Are there any placeholder values in columns?

Do more credit inquiries lead to being less likely to repay a loan?

Is there a relationship between education type, family status, or number of kids and loan repayment?

Is there a relationship between income type and loan repayment? 

Do social surroundings or region type have a relationship with loan repayment due to 
possibly indicating economic condition?

Are those with a car more likely to repay a loan?

What are the external sources and could they be a normalized version of credit scores? Would
that have a relationship with loan repayment?

Why are days birth, days employed, days registration, and days id publish negative? Does it make sense to make them positive? 

Will changing placeholder values in days employed make a big difference?

# EDA

## Package loading
```{r setup, message = FALSE, warning = FALSE}
# Set working directory
mydir <- getwd()

setwd(mydir)

library(tidyverse)
library(janitor)
library(psych)
library(skimr)
```

## Data Import
```{r data loading, message = FALSE, warning = FALSE}
# load data
application_train = read_csv("application_train.csv", col_names = TRUE, show_col_types = FALSE)
application_test = read_csv("application_test.csv", col_names = TRUE, , show_col_types = FALSE)
```

## Quick Cleaning
```{r data cleaning, warning = FALSE}
# check structure of data and ensure it loaded correctly
head(application_train)

# clean column names, make them lowercase
application_train <- application_train |>
  clean_names()

application_test <- application_test |>
  clean_names()

# factor all character variables for analysis
application_train <- application_train |>
  mutate(across(where(is.character), as.factor),
         target = as.factor(target)) 

application_test <- application_test |>
  mutate(across(where(is.character), as.factor)) 
```

## Exploratory Visualizations and Summary Tables

### Target Variable Exploration
```{r target variable exploration, warning = FALSE}
# create barplot for target variable
ggplot(application_train, aes(x = target)) +
  geom_bar() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Distribution of Target Variable",
       x = "Target (0 = No Difficulties, 1 = Payment Difficulties)",
       y = "Count") +
  theme_minimal()


# proportions of target variable for training data frame
application_train |>
  pull(target) |>
  table() |>
  prop.table() |>
  round(2)
```

The target variable is very unbalanced. 92% of the values are 0, with only 8% of the
values having troubles repaying loans. A majority class classifier then would be
92% accurate if we just wanted to predict whether or not they would have problems
repaying loans rather than predicting the probability that they would have problems 
or not. 

### Feature Exploration

```{r variable exploration, warning = FALSE}
# barplot for gender by target
application_train |> 
  ggplot(aes(x = code_gender, fill = target)) +
  geom_bar(position = "dodge") + 
  labs(title = "Distribution of Gender by Target",
       x = "Gender",
       y = "Count")
```

The barplot values look very similar, there does not seem to be any difference
in male or females having troubles repaying loans. 

```{r income, warning = FALSE}
# boxplot to see if income has effect on target
application_train |>
  ggplot(aes(x = target, y = amt_income_total)) +
  geom_boxplot() +
  labs(title = "Distribution of Income by Target",
       x = "Target",
       y = "Income")

# outlier income for 1 affecting boxplot, filter for less than 5000000 to see if that helps
application_train |>
  filter(amt_income_total < 5000000) |>
  ggplot(aes(x = target, y = amt_income_total)) +
  geom_boxplot() +
  labs(title = "Distribution of Income by Target",
       x = "Target",
       y = "Income")
```

After taking out the outlier for trouble of repaying loans, there did not seem 
to be much of a difference when it comes to income for repaying loans.

```{r credit amount, warning = FALSE}
# boxplot to compare credit amount and target
application_train |>
  ggplot(aes(x = target, y = amt_credit)) +
  geom_boxplot() +
  labs(title = "Distribution of Credit Amount by Target",
       x = "Target",
       y = "Credit Amount")
```

The two boxplots look to very similar, doesn't seem to be much of a diference 
for the target variable.

```{r good price, warning = FALSE}
# boxplot for the price of goods and target
application_train |>
  ggplot(aes(x = target, y = amt_goods_price)) +
  geom_boxplot() +
  labs(title = "Distribution of Price of Goods by Target",
       x = "Target",
       y = "Price of Goods")
```

Looks very similar, does not seem to be any differences between the target values.

```{r credit inquiries, warning = FALSE}
# boxplot to see if amount of credit inquiries varies between target values
application_train |>
  ggplot(aes(x = target, y = amt_req_credit_bureau_year)) +
  geom_boxplot() +
  labs(title = "Distribution of Amount of Credit Inquiries by Target",
       x = "Target",
       y = "Amount of Credit Inquiries")

# looks like there could be a slight difference. Lets use skim to see specific values
application_train |>
  group_by(target) |>
  skim(amt_req_credit_bureau_year)
```

There looks to be slightly more requests for those with loan payment struggles, 
but it may not be enough of a difference to be very important.

```{r days employed, warning = FALSE}

# check summary for days employed
summary(application_train$days_employed)
# 365243 days employed doesn't make sense. Possible placeholder?

# plot to see if there is large amount of these
hist(application_train$days_employed, breaks = 100, main = "Days Employed distribution", xlab = "Days Employed")
# big amount of data points at 365243, looks to be placeholder for NA

# get absolute values for negative values, change 365243 to NA
application_train <- application_train |>
  mutate(across(c(days_birth, days_employed, days_id_publish, days_registration), abs),
         days_employed = ifelse(days_employed == 365243, NA, days_employed))

## do this for testing dataset too
application_test <- application_test |>
  mutate(across(c(days_birth, days_employed, days_id_publish, days_registration), abs),
         days_employed = ifelse(days_employed == 365243, NA, days_employed))

# boxplot for days employed by target
application_train |>
  ggplot(aes(x = target, y = days_employed)) +
  geom_boxplot() + 
  labs(title = "Distribution of Days employed by Target",
       x = "Target",
       y = "Days Employed")

# seems to be slight difference between the two targets
# Looking at actual values to see difference
application_train |>
  group_by(target) |>
  skim(days_employed)
```

There was a placeholder value of 365243 for this column. I changed those placeholders
to NA to get a more accurate look at the distribution. There seems to be a slight difference in groups for days employed. Those with loan repayment problems may be employed for smaller amounts of time.

```{r days birth, warning = FALSE}
# check days birth to see if there are any outliers
summary(application_train$days_birth) 

# create boxplot for days birth and target
# dividing by 365 to make it more readable in years
application_train |>
  ggplot(aes(x = target, y = days_birth / 365)) +
  geom_boxplot() +
  labs(title = "Distribution of Age by Target",
       x = "Target",
       y = "Age")

# Looks to be possible difference
application_train |>
  group_by(target) |>
  skim(days_birth)
```

There seems to be a bit of a difference in age and loan repayment difficulties. 
It looks like those who are older do not have as many loan repayment problems. 

```{r identity document, warning = FALSE}
# quick check to see if days id publish have any placeholders
# create boxplot for identity document dayas by target
summary(application_train$days_id_publish)
application_train |>
  ggplot(aes(x = target, y = days_id_publish)) +
  geom_boxplot() +
  labs(title = "Days before application that client changed identity document by Target",
       x = "Target",
       y = "Days")
```

There is a small difference between the two, but doesn't look to be significant.

```{r registration, warning = FALSE}
# quick check to see if days registration have any placeholders
summary(application_train$days_registration)
# create boxplot for registration days by target
application_train |>
  ggplot(aes(x = target, y = days_registration)) +
  geom_boxplot() +
  labs(title = "Days before application that client changed registration by Target",
       x = "Target",
       y = "Days")
```

There does not look to be any difference for the target values in days before 
the application that the client changed their registration

```{r income type, warning = FALSE, fig.width = 10}
# barplot to see if income type matters for target
application_train |> 
  ggplot(aes(x = name_income_type, fill = target)) +
  geom_bar(position = "fill") + # doing fill to see propotion rather than raw counts
  labs(title = "Income Type by Target",
       x = "Income Type",
       y = "Count")
# high proportion of problems for unemployed and maternity leave, see if counts for those professions are large enough to be something
table(application_train$name_income_type)
```

When initially looking at the proportions for income type by target, unemployed and
maternity leave had high proportions. However, there were only 5 values for maternity
leave ad 22 for unemployed, which does not seem to be big enough values to matter.
Overall, there does not look to be a difference in income type and loan repayment. 

```{r education type, warning = FALSE, fig.width = 10}
# barplot for education type by target
application_train |> 
  ggplot(aes(x = name_education_type, fill = target)) +
  geom_bar(position = "fill") + # using fill again for proportions rather than counts
  labs(title = "Education Type by Target",
       x = "Education Type",
       y = "Count") 
```

Lower Secondary has a slightly higher proportion of loan payment difficulties than
the the others, but doesn't look to be significant enough amount to be a big difference

```{r family status, warning = FALSE}
# barplot for family status
application_train |>
  ggplot(aes(x = name_family_status, fill = target)) +
  geom_bar(position = "fill") + 
  labs(title = "Family Status by Target",
       x = "Family Status",
       y = "Count")
```

The proportions for family status and the target are all very similar. Does not look
to be important in determining loan repayment difficulties.

```{r number of kids, warning = FALSE}
# use skim to see amount of children by target
application_train |>
  group_by(target) |>
  skim(cnt_children)
```

The number of children for both values of targets are very sinilar and thus there
does not seem to be a difference.

```{r external scores, warning = FALSE}
# Using skim to see numbers for external sources by target
application_train |>
  group_by(target) |>
  skim(ext_source_1)
# does look to be lower score for 1, lots of missing values

# does that fit for 2 and 3?
application_train |>
  group_by(target) |>
  skim(ext_source_2)

# bigger sample size, looks to be slightly lower score for 1
application_train |>
  group_by(target) |>
  skim(ext_source_3)
# looks to hold for all three scores, 1 and 3 both missing good amount of values
```

While I do not know exactly what the external sources are, I think it's possible
they are normalized credit scores. Whatever they are, these seem to have the 
biggest differences between having loan repayment problems and not. These will be
important in determining loan payment problems.

```{r region rating, warning = FALSE}
# barplot for region rating by target
application_train |>
  ggplot(aes(x = factor(region_rating_client), fill = target)) +
  geom_bar(position = "fill") + 
  labs(title = "Region Rating by Target",
       x = "Region Rating",
       y = "Count")
# slightly higher proportion of 1's for rating of 3, but also lower count of them
```

The proportion of people who have a region rating of 3 and trouble repaying loans
is slightly higher than the others. It is possible there is a small effect on it.

```{r social surroundings, warning = FALSE}
# social surroundings by target
application_train |>
  group_by(target) |>
  skim(obs_30_cnt_social_circle)
```

The social surrounds of those defaulting are about the same those who are not defaulting.

```{r car, warning = FALSE}
# barplot for owning a car by target
application_train |> 
  ggplot(aes(x = flag_own_car, fill = target)) +
  geom_bar(position = "fill") + 
  labs(title = "Owning Car by Target",
       x = "Owning Car",
       y = "Count")

# Seeing if the age of the car changes anything
application_train |>
  group_by(target) |>
  skim(own_car_age)

application_train |>
  ggplot(aes(x = target, y = own_car_age)) +
  geom_boxplot() 
```

Neither owning a car nor the age of a car looks to be any different when it comes
to repaying loans.

```{r correlation, warning = FALSE}
# seeing if there is any correlation between the external sources
#creating train copy to make target variable numeric to find correlation
train_copy <- application_train
train_copy$target = as.numeric(train_copy$target)

# finding correlation between target and the external sources
cor_result <- cor(train_copy[c("target", "ext_source_1", "ext_source_2", "ext_source_3")], use = "pairwise.complete.obs") %>% round(2)
cor_result
```

There is a slight negative correlation between the three external sources and the
target. It is small, but there is something there. 

## Missing Data
```{r missing values, warning = FALSE}
# how many total NA's?
sum(is.na(application_train))
#percent of NA's in dataset
sum(is.na(application_train)) / (ncol(application_train) * nrow(application_train)) * 100
# how many rows contain at least 1 NA?
sum(rowSums(is.na(application_train)) > 0) 
# only 8000 don't have at least 1 NA
## are there any rows that have more than 25% of NA?
sum(rowSums(is.na(application_train)) >= 30) 
## more than half have at least 25% of NA's

# percent of NA's in each column
na_percentage <- round(sapply(application_train, function(x) sum(is.na(x)) / length(x) * 100),4)
# sort percentages
na_percentage_sorted <- sort(na_percentage, decreasing = TRUE)
# show percent of columns that have more than 50% of the data missing
na_percentage_sorted[na_percentage_sorted > 50]
```

There are a lot of missing values in this dataset. Over half the rows in the dataset
have at least 25% of their values as missing. 41 columns out of the 122 columns
have at least 50% or more of their data missing, with many of these columns having to 
do with the living area spaces. I believe any columns that have over 50 or 60% of
the data missing could most likely be removed before modeling in order to improve the model.
For the rest of the missing values, I plan on using KNN to impute those values.

# Results
I think the biggest thing I have learned through EDA is that the external values, age,
and number of days employed seem to be the predictors that differ the most when
it comes to problems repaying a loan or not. These seem to be the strongest relationships, 
so when modeling, I expect those to be some of the biggest factors in determining
the probability that someone will have problems repaying their loan. Most of the
other relationships seemed to be fairly normal and not different when compared to
being able to repay a loan or not. I encountered a few data problems I encountered. 
The first is that four columns were all negative to indicate time passed. I changed
them to be all positive to make it more understandable. There was also a placeholder
value for days employed that I had to change into NAs. 

This EDA first made me realize I will have to deal with class imbalance when
modeling. I will have to look into how to do that and what the best metrics are
to evaluate modeling class imbalance. I now have an idea of what some of the most
important features are likely to be while also having a plan on which columns
I can remove due to missing values. EDA has also helped me decide that I want to
impute the missing values, most likely through KNN. Lastly, because there seems
to be complex relationships between these variables and being able to repay a loan,
I think a slightly more complex model, like a Random Forest model, would be the best
option to capture those relationships and get better results. 